Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-10-29 ms/ms.qmd
revised ms sent to coauthors

Amphistomatous leaves should be common because, all else being equal, a leaf with a given number of stomata per area could increase its photosynthetic rate simply by apportioning approximately half its stomata to each surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @gutschick_photosynthesis_1984-1]. The key difference between a hypo- and amphistomatous leaf, holding all other factors constant, is that an amphistomatous leaf has two parallel diffusion paths through the internal airspace to any given chloroplast. Those airspaces pose a resistance for CO$_2$ diffusion, so CO$_2$ concentration drops as it approaches chloroplasts. Shorter pathways mean a smaller drop in CO$_2$ concentration. Thus, chloroplasts in amphistomatous leaves experience higher CO$_2$ concentrations than in hypostomatous leaves, thereby increasing photosynthesis. The airspace resistance (or its inverse, the airspace conductance, $g_\mathrm{ias}$) is rarely measured directly and there is disagreement between empirical [@parkhurst_intercellular_1990; @morison_lateral_2005; @evans_resistances_2009; @tomas_importance_2013; @earles_beyond_2018; @santrucek_partitioning_2019; @nobel_physicochemical_2020; @harwood_understanding_2021; @marquez_assessing_2023] and theoretical models [@tholen_mechanistic_2011; @ho_three-dimensional_2016; @theroux-rancourt_maximum_2021]. The $g_\mathrm{ias}$ in thin, porous leaves may be so large as to be inconsequential given much lower conductances for other components of the diffusion pathway, whereas the $g_\mathrm{ias}$ of thick leaves with little airspace may greatly hinder CO$_2$ diffusion to chloroplasts. Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$. The airspace conductance is one component of the overall mesophyll conductance, $g_\mathrm{m}$, which is often strongly influenced by the chloroplast surface area exposed to airspace and mesophyll cell wall thickness [@evans_resistances_2009; @gago_photosynthesis_2020; @flexas_cell_2021]. Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but would still benefit from amphistomy.

Amphistomy should also enhance photosynthesis when leaf boundary layer resistance is high, because apportioning total flux between two boundary layers rather than one results in a smaller CO$_2$ concentration drop between the atmosphere and stomata. A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019]. However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2021] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.

@@

We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification. For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values (see below for details). We refer to this as an $A \textendash g_\text{sw}$ curve. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm). Hypostomy refers to leaves with stomata only present on the lower, typically abaxial, surface. We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

@@

The log-response ratio is commonly used social and biological sciences (e.g. @hedges_metaanalysis_1999). It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

@@

It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$. To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$. In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\mathrm{AA}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. We fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023]. We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023]. We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations per chain. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$, meaning the 95\% confidence intervals of $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}$ should be positive and not encompass 0.

@@

We estimated the relationship between leaf $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ measured from a single leaf per population. We propagated uncertainty about $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$.

@@

This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common. By developing a straightforward experimental method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant. In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest. This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that be an important consideration in predicting realized photosynthetic rates in nature. Rather, our experiments show that coastal amphistomatous leaves with the same total leaf stomatal conductance photosynthesize more than identical hypostomatous leaves. We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.

>>>>

Amphistomatous leaves should be common because, all else being equal, a leaf with a given number of stomata per area could increase its photosynthetic rate simply by apportioning approximately half its stomata to each surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @gutschick_photosynthesis_1984-1]. The key difference between a hypo- and amphistomatous leaf, holding all other factors constant, is that an amphistomatous leaf has two parallel diffusion paths through the internal airspace to any given chloroplast. Those airspaces pose a resistance for CO$_2$ diffusion, so CO$_2$ concentration drops as it approaches chloroplasts. Shorter pathways mean a smaller drop in CO$_2$ concentration. Thus, chloroplasts in amphistomatous leaves experience higher CO$_2$ concentrations than in hypostomatous leaves, thereby increasing photosynthesis. The airspace resistance (or its inverse, the airspace conductance, $g_\mathrm{ias}$) is rarely measured directly and there is disagreement between empirical [@parkhurst_intercellular_1990; @morison_lateral_2005; @evans_resistances_2009; @tomas_importance_2013; @earles_beyond_2018; @santrucek_partitioning_2019; @nobel_physicochemical_2020; @harwood_understanding_2021; @marquez_assessing_2023] and theoretical models [@tholen_mechanistic_2011; @ho_three-dimensional_2016; @theroux-rancourt_maximum_2021]. The $g_\mathrm{ias}$ in thin, porous leaves may be so large as to be inconsequential given much lower conductances for other components of the diffusion pathway, whereas the $g_\mathrm{ias}$ of thick leaves with little airspace may greatly hinder CO$_2$ diffusion to chloroplasts. Amphistomy should confer the largest photosynthetic benefit in leaves with intrinsically low $g_\mathrm{ias}$. The airspace conductance is one component of the overall mesophyll conductance, $g_\mathrm{m}$, which is often strongly influenced by the chloroplast surface area exposed to airspace and mesophyll cell wall thickness [@evans_resistances_2009; @gago_photosynthesis_2020; @flexas_cell_2021]. Hence, thicker leaves may compensate for lower $g_\mathrm{ias}$ through increased chloroplast surface area exposed to airspace [@terashima_irradiance_2006], but will still benefit from amphistomy as long as $g_\mathrm{ias}$ is finite.

Amphistomy should also enhance photosynthesis when leaf boundary layer resistance is high, because apportioning total flux between two boundary layers rather than one results in a smaller CO$_2$ concentration drop between the atmosphere and stomata. A similar effect has been validated with a computer model and measurements for transpiration: amphistomatous leaves lose somewhat more water for the same vapor pressure deficit and total $g_\mathrm{sw}$ [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost under most realistic conditions [@muir_is_2019]. However, if minimal stomatal conductance is related to stomatal density [@drake_smaller_2013; @marquez_cuticular_2022] and the upper boundary layer conductance is higher, then amphistomy could cause additional, unavoidable water loss.

@@

We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification. For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values. We refer to this as an $A \textendash g_\text{sw}$ curve, which is described in more detail below. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm). Hypostomy refers to leaves with stomata only present on the lower, typically abaxial, surface. We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

@@

The log-response ratio is commonly used social and biological sciences (e.g. @hedges_meta-analysis_1999). It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The log-response ratio is preferable to the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

@@

It is not feasible to record $A_\mathrm{amphi}$ and $A_\mathrm{hypo}$ at the exact same $g_\text{sw}$. To overcome this, we fit $A \textendash g_\text{sw}$ curves using a linear regression of $\text{log}(g_\mathrm{sw})$ on $A$ to interpolate modeled $A$ for amphi and pseudohypo leaves at the same $g_\text{sw}$. Let $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ be the estimated $A$ of the amphi and pseudohypo leaves, respectively. We estimated these quantities at the same $g_\mathrm{sw}$ using fitted parameters ($\hat{\beta}$'s):

$$\hat{A}_\text{amphi} = \hat{\beta}_{0,\text{amphi}} + \hat{\beta}_{1,\text{amphi}} \times \text{log}(g_\mathrm{sw})$$
$$\hat{A}_\text{hypo} = \hat{\beta}_{0,\text{hypo}} + \hat{\beta}_{1,\text{hypo}} \times \text{log}(g_\mathrm{sw})$$

In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf. $\mathrm{AA}$ for each leaf was calculated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ using the log-response ratio shown above.

To estimate $\hat{\beta}$'s from the $A \textendash g_\text{sw}$ curve for each leaf, we fit Bayesian regressions using the *R* package **brms** version `r fit_licor$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023]. We used CmdStan version `r fit_licor$version$cmdstan` and **cmdstanr** version `r fit_licor$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023]. We sampled the posterior distribution from `r dim_fit_licor[2]` chains with `r dim_fit_licor[1]` iterations each after `r dim_fit_licor[1]` warmup iterations per chain. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. The key prediction is that $\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}$, meaning the 95\% confidence intervals of $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}$ should be positive and not encompass 0.

@@

We estimated the relationship between leaf $g_\mathrm{smax,ratio}$ and $\mathrm{AA}$ measured from a single leaf per population. We propagated uncertainty about $\mathrm{AA}$ and $g_\mathrm{smax,ratio}$ by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of $g_\mathrm{smax,ratio}$ on $\mathrm{AA}$ is positive until $g_\mathrm{smax,ratio} < 0.5$, meaning the 95\% confidence interval of the slope should be positive and not encompass 0 in the domain $g_\mathrm{smax,ratio} < 0.5$.

@@

This study reports the first direct experimental evidence that having stomata open on both leaf surfaces, amphistomy, increases photosynthesis for a given total stomatal conductance, particularly in leaves from the type of open, sunny habitats where this trait is most common. By developing a straightforward experimental method to block gas exchange through the upper surface, we directly compared the photosynthetic rate of a leaf with gas exchange through both surfaces or just one, holding all other factors constant. In doing so, we found that coastal leaves of the indigenous Hawaiian ʻilima (*Sida fallax*) enjoyed a greater photosynthetic benefit from amphistomy than nearby montane leaves living in more closed forest. This is not because amphistomatous leaves necessarily have greater leaf surface available for stomata, although that likely influences realized photosynthetic rates in natural populations. Rather, our experiments show that coastal amphistomatous leaves with the same total leaf stomatal conductance photosynthesize more than identical hypostomatous leaves. We cannot yet ascribe the difference in amphistomy advantage between coastal and montane leaves to particular physiological or anatomical variation, but this is a promising area for future research.
